Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-04 Manuscript/manuscript.Rmd
manuscript edits

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. For example, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeat for each subsequent resolution.

@@

In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling), composed of data from each predictor type. Each re-sampling technique is described in detail below.

@@

A regularization technique using a combination of the $l_{1}$ and $l_{2}$ penalties, known as the elastic-net, was used to reduce the feature space. The elastic-net was chosen in order to suitably reduce the feature space while also maintaining the inclusion of multiple correlated predictors. For each elastic-net regularization, 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. All elastic-net algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret].

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to fully compare re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used <!--to tune the number of features to consider at each node and the number of ensemble trees to aggregate-->. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was assessed using the test set, and performance metrics were recorded.

The efficacy of each ensemble system in the framework was compared using composite metrics including F1-score and MCC. These metrics are defined as follows:

$$F1 - Score = \frac{2TP}{2TP + FP + FN}$$

@@

## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE AMONG ALL CLASSES OF MODELS

@@

## DATA RE-SAMPLING COUPLED WITH THE INCLUSION OF A DISTANCE-TYPE FEATURE SPACE IMPROVED PERFORMANCES AMONG ALL CLASSES OF MODELS

>>>>

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0). The process was repeated for each subsequent resolution.

@@

In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling), composed of the genomic bin class and data from each predictor type. The re-sampling techniques considered are described in detail below.

@@

A regularization technique using a combination of the $l_{1}$ and $l_{2}$ penalties, known as the elastic-net, was used to reduce the feature space. The elastic-net was chosen in order to suitably reduce the feature space while also maintaining the inclusion of multiple correlated predictors. For each elastic-net regularization, 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. The parameters that maximized the mean AUPRC across each cross-fold was chosen. All elastic-net algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret].

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to compare model performances between re-sampling techniques and predictor types. <!--To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features to consider at each node and the number of ensemble trees to aggregate-->. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set.

The performance of each ensemble system in the framework was compared using the composite metric, Matthew's Correlation Coefficient. The MCC is defined as follows:

<!--
$$F1 - Score = \frac{2TP}{2TP + FP + FN}$$
-->

@@

## CLASS IMBALANCE SEVERELY IMPACTS MODEL PERFORMANCE

@@

## DATA RE-SAMPLING COUPLED WITH THE INCLUSION OF A DISTANCE-TYPE FEATURE SPACE SUBSTANTIALLY IMPROVED MODEL PERFORMANCE
